The purpose of this script is to organize and clean up the Bighorn dataset. Huge credit to Laura Eliuk for leaving this project in such great shape. 

Main outputs we want
1. List of raw detections
2. List of independent detections to 30 minutes threshold
3. Camera deployment file
4. Site covariate file

Andrew Barnas 
June 12th 2025
andrew.f.barnas@gmail.com

Load in packages we will need, setup working directories, and load in files. I think this should be fairly straightforward since Laura left this really well organized. 
```{r}
#Clear everything out and start fresh
rm(list=ls())

library(tidyr)        # data tidying
library(stringr)      # working with strings
library(dplyr)        # data manipulation
library(reproducible) # reproducible working directories
library(ggplot2)      # plots
library(ggpubr)       # combining plots
library(lubridate)   # dates


#Setup input and output paths
input_directory <- checkPath(file.path(getwd(), "inputs"), 
                            create = TRUE)

output_directory <- checkPath(file.path(getwd(), "outputs"), 
                             create = TRUE)


#Read in the detection file
dets<-read.csv(file.path(input_directory, "BH.detections.raw.csv"), header = TRUE)

#Read in the covariate file
covs<-read.csv(file.path(input_directory, "bh.landcover.all.csv"), header = TRUE)

#Read in the deployment file
camop<-read.csv(file.path(input_directory, "BH_deployment_clean.csv"), header = TRUE)



```

Ok lets do our initial checking. Now Lauras thesis states there were 93 cameras deployed but only 83 functional sites. 
```{r}

#What do we have in the deteciton file? Looks like pretty standard timelapse files
colnames(dets)

#Bless you laura, there are 83 sites.
n_distinct(dets$Site)


#Ok now what about the covariates file? 92, but maybe that was before filting and stuff
n_distinct(covs$Site)
#Are all the sites in the detection file in the covariates file?

unique(dets$Site)
unique(covs$Site)

#Nice, looks like everyone is accounted for so thats sicks
dets%>%
  filter(!Site %in% unique(covs$Site))

#What about deployment? Do we have 83 sites? No 93, but again who cares we have data for all
n_distinct(camop$Deployment.Location.ID)
unique(camop$Deployment.Location.ID)
dets%>%
  filter(!Site %in% unique(camop$Deployment.Location.ID))



```

Alright well...I guess lets get these all on the same page
```{r}
#First some simple renaming 
dets<-dets%>%rename(site = Site)
covs<-covs%>%rename(site = Site)
camop<-camop%>%rename(site = Deployment.Location.ID)

#And now lets filter the covariate and camop dataframe to only have those 83 sites
camop<-camop%>%
  filter(site %in% unique(dets$site))

covs<-covs%>%
  filter(site %in% unique(dets$site))

```

Formating the camop file
```{r}
camop<-camop%>%
  select(site,
         Latitude,
         Longitude,
         start_date = Camera.Deployment.Begin.Date,
         end_date = Camera.Deployment.End.Date)

#Quick ganntt chart to see if it makes sense. A few weird cameras starting in spring 2020, but I guess not out of the ordinary?
camop%>%
  mutate(start_date = mdy(start_date),
         end_date = dmy(end_date))%>%
   ggplot(aes(y = site))+
  geom_segment(aes(x = start_date, xend = end_date, y = site, yend = site))


#And i guess thats it, lets write the file
write.csv(camop, file.path(output_directory, "bighorn_camop.csv"), row.names = FALSE)

```

Formatting the detection files. The purpose here is to present the raw detection data, and then the standard 30 minute independent detection file.
```{r}
#Cleaning up the detection file here slightly
dets<-dets%>%
  select(site, 
         datetime = Date.Time,
         species = Species)

#Alright curious what we have in here
dets%>%
  group_by(species)%>%
  summarise(n())

#Ok so there are obviously some classification issues here sort of. Like snowmobiler vs human, a bunch of unknown misc species
unique(dets$species)

#So here is the strategy. I am just going to output this as the raw detection file because thats what it is. 
write.csv(dets, file.path(output_directory, "bighorn_detections_raw.csv"), row.names = FALSE)


#But then do a bit of cleaning up for the 30 minute detection file. I will remove the empty images, and all "unknown" species. I think that is a fair approach. If anyone wants to get more specific, they can do it themselves with the raw data!
dets<-dets%>%
  #Remove blanks
  filter(species != "")%>%
  #Get rid of all those unknown species
  filter(!str_detect(species, 'Unknown'))%>%
  #There is also an "Other birds"....which is basically unknown
  filter(!str_detect(species, 'Other'))
  
#That worked nicely
unique(dets$species)
#Nice, still have 83 sites
n_distinct(dets$site)


#Ok lets run the independent deteciton loop
dets<-dets%>%
  arrange(site, species, datetime)%>%
  group_by(site, species)%>%
  mutate(timediff = as.numeric(difftime(datetime, lag(datetime),
                                        units = "mins")))

#This is our rolling window for "independence": we will count windows as independent if they are more than "mins" apart. Here we are using 30 minutes. 

mins<-30

#This loop assigns group ID based on independence. 
 dets$Event.ID <- 9999
 seq <- as.numeric(paste0(nrow(dets),0))
 seq <- round(seq,-(nchar(seq)))
 for (i in 2:nrow(dets)) {
   dets$Event.ID[i-1]  <- paste0("E",format(seq, scientific = F))
   if(is.na(dets$timediff[i]) | abs(dets$timediff[i]) > (mins)){
     seq <- seq + 1
   }
 }
 
 if(dets$timediff[nrow(dets)] < (mins)|
    is.na(dets$timediff[nrow(dets)])){
   dets$Event.ID[nrow(dets)] <- dets$Event.ID[nrow(dets)-1]
 } else{dets$Event.ID[nrow(dets)] <- paste0("E",format(seq+1, scientific = F))
 }
 
   #And then take only the first row of data for each independent event
dets_independent<-dets%>%
   group_by(Event.ID)%>%
   slice(1)%>%
  arrange(site, species)

#And write the independent detection file
write.csv(dets_independent, file.path(output_directory, "bighorn_30min_independent_detections.csv"),
          row.names = FALSE)

```

The only thing that needs to be done to the covariate file is remove one silly column
```{r}
covs<-covs%>%
  select(-Array)

#Write the covariate file
write.csv(covs, file.path(output_directory, "bighorn_covariates.csv"),
          row.names = FALSE)

```



Fin.